Binning Strategies and Related Loss for Large Data

### 2015-03-01 Paper.Rnw
Edits to Section 2

\section{Scatterplots for Large Data Sets}
In the case of medium sized data, scatterplots are great tools for showing relationships in two dimensions. For large data, scatterplots suffer from over-plotting -- i.e. more and more points are drawn in close-by places, thereby masking relevant structure.
Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted versus number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
Figure~\ref{fig:scatter-alpha} shows a traditional scatterplot on the left. Each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts. Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points visible that are close together. This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points. 
%\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.

@@

All of these methods fall short in the example.
As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information. Tukey's open circles (b) help \hh{to a degree, but are also prone to over-plotting  when the data set is very large.} Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers. The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.

@@

\ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales. \hh{Address how ties in the data are dealt with, which are very common for large data sets due to the finite resolution of data collection (e.g.\ airline data is collected at minute intervals). jittering?}  Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points. A\hh{n} extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into many segments on different scales. Theses approaches make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.} \hh{please check the spelling in the previous paragraph. }

@@

Histograms are an example of plots which use binned variables. \hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information. Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.}
%To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step
\hh{A natural extensions of histograms to higher dimensions is} to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, \hh{similar to} a two-dimensional histogram viewed from above.

@@

When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.  When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion. \hh{However,} these estimates require \hh{careful} parameter tuning, \hh{as over-smoothing} may hide gaps in the data while simultaneously de-emphasizes outlying points.

>>>>

\section{Scatterplots for Large Data Sets}
In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions. For large data, scatterplots suffer from over-plotting -- i.e. \ktm{a large number of observations drawn in such close proximity that their points overlap}, thereby masking relevant structure.
Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted versus number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
Figure~\ref{fig:scatter-alpha} shows a traditional scatterplot on the left. Each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts. Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points visible that are close together. This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points. 
%\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.

@@

All of these methods fall short in the example.
As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information. Tukey's open circles (b) help to a degree, but are also prone to over-plotting  when the data set is very large. Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers. The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.

@@

\ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales. Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations by having points repel one another to avoid overlapping points.  An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into many segments on different scales and also does not deal with points that are exactly overlapped due to the resolution of data collection. Generalized and variable-binned scatterplots make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.}

@@

%Histograms are an example of plots which use binned variables. \hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information. Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.} 
\ktm{Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.}  A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, similar to a two-dimensional histogram viewed from above.

@@

When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.  When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion. However, these estimates require careful parameter tuning, as over-smoothing may hide gaps in the data while simultaneously de-emphasizes outlying points.
